---
title: "Population of Kanton Zürich with statR and OGD data"
format:   
  html:
    embed-resources: true
editor: visual
---

## Testing the statR package

This is a small analysis to familiarize myself with some of the functionalities of the statR package as well as the OGD data of the statistical office of the canton of Zurich. I visualize the 2025 population data of the municipalities of the canton of Zurich as a map.  

```{r}
#| eval: false

#install statR from GitHub (only once)
install.packages("downloader")
library(downloader)

download("https://github.com/statistikZH/statR/archive/refs/heads/master.tar.gz", "statR.tar.gz")
install.packages("statR.tar.gz", repos = NULL, type = "source")

```

```{r}
#| message: false
# install relevant packages 
# needed only once, add all packages you do not have installed
# install.packages(c("sf", "tidyverse","scales")) 

#load packages
library(sf)
library(tidyverse)
library(statR)
library(scales)

```

```{r}
#| label: load-geodata-municipalities
#| message: false
#| warning: false

# Load shape file for the "Gemeindegrenzen" of the canton Zurich

# create directory and load shape data. I use the 2020 shape files like in the statR documentation. 
dir.create("data", showWarnings = FALSE)
if (!dir.exists("data/GEN_A4_GEMEINDEN_2020_shp")) {
  download.file(
    "https://www.web.statistik.zh.ch/ogd/daten/ressourcen/KTZH_00000151_00001653.zip",
    "data/geodata.zip", mode = "wb"        
  )
  unzip("data/geodata.zip", exdir = "data")
}

muni <- st_read("data/GEN_A4_GEMEINDEN_2020_shp", quiet = TRUE)

#check names
names(muni)  
```

```{r}
#| label: load-population-data
#| message: false

# Population data from the OGD catalouge of the canton zurich 
pop_url <- "https://www.web.statistik.zh.ch/ogd/data/KANTON_ZUERICH_133.csv"

#read population data
pop_raw <- read.csv(pop_url, fileEncoding = "UTF-8")
names(pop_raw) #check column names

#check latest available year
year <- max(pop_raw$INDIKATOR_JAHR, na.rm = TRUE)
year
```

```{r}
#| label: prepare-population-data

# aggregate population  data for the latest year available 
pop_gem <- pop_raw %>%
  filter(INDIKATOR_JAHR == year) %>%
  group_by(bfs = as.character(BFS_NR)) %>%
  summarise(bevoelkerung = sum(INDIKATOR_VALUE, na.rm = TRUE), .groups = "drop")
```

```{r}
#| label: join-data

# Join population data with the geodata
gemeinden_pop <- muni%>%
  mutate(bfs = as.character(BFS)) %>%
  left_join(pop_gem, by = "bfs")
```

```{r}
#| label: color-palette-all
#check all available scales to select a nice one for the visualisations
display.statcol.all()  
```

```{r}
#| label: color-palette-blue

#check the available zueri blue color grades in the statR package
show_col(zhpal$zhblue)
```

```{r}
#| label: histogram-bevoelkerung
#| message: false
#| warning: false

# Summary statistics of the population variable
summary(gemeinden_pop$bevoelkerung)

# plot the distribution of the population variable
barfill <- "#3F8AB4" # use second darkest zueri blue
ggplot(gemeinden_pop %>% filter(!is.na(bevoelkerung)),
       aes(x = bevoelkerung)) +
  geom_histogram(bins = 60, fill = barfill, color = "white") +
  scale_x_continuous(labels = scales::label_number(big.mark = "'")) +
  labs(
    title    = "Verteilung der Gemeindebevölkerung 2025",
    subtitle = "Kanton Zürich – Anzahl Gemeinden je Grössenklasse",
    x = "Einwohner pro Gemeinde", y = "Anzahl Gemeinden",
    caption  = "Quelle: Statistisches Amt Kanton Zürich (OGD)"
  ) +
  theme_stat()
```

-\> Data is extremely right skewed. Log scale might be better for the visualisation.

```{r}
#| label: histogram-bevoelkerung-log
#| message: false
#| warning: false

#check the distribution of the population variable on log scale
ggplot(gemeinden_pop %>% filter(!is.na(bevoelkerung)),
       aes(x = bevoelkerung)) +
  geom_histogram(bins = 60, fill = barfill, color = "white") +
  scale_x_log10(labels = scales::label_number(big.mark = "'")) +
  labs(
    title    = "Verteilung der Gemeindebevölkerung 2025 (log-Skala)",
    subtitle = "Kanton Zürich",
    x = "Einwohner pro Gemeinde (log10)", y = "Anzahl Gemeinden"
  ) +
  theme_stat()
```

-\> distribution is now much more normal, better for the visualisation

```{r}
#| label: map-canton-default
#| fig-width: 7
#| fig-height: 8
#| message: false
#| warning: false

# plot map of the canton Zurich with the population data 
ggplot(gemeinden_pop, aes(fill = bevoelkerung)) +
  geom_sf(color = "white", linewidth = 0.1) +
  scale_fill_gradientn(
    colours  = rev(zhpal$zhblue), # zueri blue for color sclae 
    name     = "Personen",
    trans    = "log10", # transform scale, otherwise zuerich and winterthur dominate the color scale
    breaks   = c(1000, 5000, 20000, 100000, max(gemeinden_pop$bevoelkerung, na.rm = TRUE)),     # max as upper limit
    labels   = label_number(big.mark = "'"),
    na.value = "#D8D2C8",
    guide    = guide_colourbar(barheight = grid::unit(8, "cm")) 
  ) +
  labs(
    title    = "Bevölkerung im Kanton Zürich (2025)",
    subtitle = "nach Gemeinde · Farbskala logarithmisch (log10)",
    caption  = "Quelle: Statistisches Amt Kanton Zürich (OGD) · Geodaten: GIS-ZH (Gemeinden 2020)"
  ) +
  theme_stat(map = TRUE)
```

```{r}
#| label: map-canton-vibrant
#| fig-width: 7
#| fig-height: 8
#| message: false
#| warning: false

# plot map of the canton Zurich with a more vibrant color scale
ggplot(gemeinden_pop, aes(fill = bevoelkerung)) +
  geom_sf(color = "white", linewidth = 0.1) +
  scale_fill_gradientn(
    colours  = rev(zhpal$zhbyg), # zhbyg color scale 
    name     = "Personen",
    trans    = "log10", 
    breaks   = c(1000, 5000, 20000, 100000, max(gemeinden_pop$bevoelkerung, na.rm = TRUE)),     # max as upper limit
    labels   = label_number(big.mark = "'"),
    na.value = "#D8D2C8",
    guide    = guide_colourbar(barheight = grid::unit(8, "cm")) 
  ) +
  labs(
    title    = "Bevölkerung im Kanton Zürich (2025)",
    subtitle = "nach Gemeinde · Farbskala logarithmisch (log10)",
    caption  = "Quelle: Statistisches Amt Kanton Zürich (OGD) · Geodaten: GIS-ZH (Gemeinden 2020)"
  ) +
  theme_stat(map = TRUE)
```

-\> vibrant color scale is less elegant but does make the difference between similarly sized municipalities more immediately visible.
